Guidelines for Submissions to ISCA 2013

### 2012-11-18 paper/isca40format/stuff/AbstractProtocol.tex
stuff

\newcommand{\One}[2]{\text{#1$\langle#2\rangle$}}
\newcommand{\two}[3]{\text{#1$\langle#2,#3\rangle$}}
\newcommand{\three}[4]{\text{#1$\langle#2,#3,#4\rangle$}}
\newcommand{\four}[5]{\text{#1$\langle#2,#3,#4,#5\rangle$}}
\newcommand{\True}{\texttt{True}}
\newcommand{\False}{\texttt{False}}
\newcommand{\assigns}{:=}
\newcommand{\AndAnd}{\;\&\&\;}
\newcommand{\OrOr}{\;||\;}
\newcommand{\Inv}{\text{$Inv$}}

@@

The notation we adopt is primarily for conciseness of description while allowing
maximum flexibility in the implementation.

A node keeps information about the state of each cache line in the node as well
as in each of its children, \ie the directory. Figure~\ref{table:states} shows
the precise information that a node keeps for each cache line in the node. The
notation we adopt is primarily for conciseness of description. For instance, the
directory need not contain the exact copy of each of the child's coherence
state, for each cache line, instead the encoding can be more efficient. We will
show efficient implementations later.

\begin{figure}
\begin{tabular}{|l|p{.64\columnwidth}|}
\hline
isPresent(a) & Check if cache line for address $a$ is present\\
\hline
data(a) & Cache line data for address $a$\\
\hline
state(a) & Coherence state for address $a$\\
\hline
dir(a, c) & Directory's copy of the coherence state for address $a$ in child
$c$\\
\hline
isCWait(c, a) & Check if the node waiting for a response from child $c$
for address $a$\\
\hline
cWait(c, a) & The coherence state to which the node has requested child $c$ to
go to, for address $a$\\
\hline
isPWait(a) & Check if the node waiting for a response from its parent for
address $a$\\
\hline
whoTrigger(a) & The source of the request which made the node wait for a
response. It is $None, Parent, Child \text{ or } Evict$\\
\hline
replacingAddr(a) & If $whoTrigger(a)$ is $Evict$, the address replacing the line
of address $a$\\
\hline
childTrigger(a) & If $whoTrigger(a)$ is $Child$, the index of the child which
is the source of the request\\
\hline
\end{tabular}
\caption{Information that the node keeps for a cache line}
\label{table:states}
\end{figure}

The values of the coherence state (and hence the directory state) belong to a
totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
defined between every pair of elements in that set and that relation is
transitive. For instance, in an MSI protocol, the values of the coherence state
are from the set $\{M, S, I\}$ where $I < S$, $S < M$ and $I < M$. Our abstract
protocol does not assign any semantics or meaning to the values of the coherence
states. In a concrete cache coherence protocol, the values of the states are
associated with permissions for reading and writing, and the state with fewer
permissions is $<$ the state with more permissions. We use \Inv to denote the
least element in the set containing the values of the coherence state.
Typically, this denotes that the node has no permissions for reading or writing
that line.

@@

Figure~\ref{table:messages} shows the incoming and outgoing messages to and from
the node, respectively, and requests for eviction. An eviction request for an
address is generated if a child has made a request for a different cache line
whose address is not currently present in the cache.

\begin{figure}
\begin{tabular}{|l|p{.6\columnwidth}|}
\hline
\multicolumn{2}{|c|}{Incoming messages}\\
\hline
\three{ReqFromC}{c}{a}{x} & Request from child $c$ upgrade $c$'s coherence
state for address $a$ to $x$\\
\two{ReqFromP}{a}{x} & Request from parent to downgrade node's coherence
state for address $a$ to $x$\\
\three{RespFromC}{c}{a}{x} & Message from child $c$ denoting that $c$'s
coherence state for address $a$ has changed to $x$\\
\three{DataFromC}{c}{a}{d} & Data $d$ of the cache line received from $c$ for
address $a$\\
\two{RespFromP}{a}{x} & Response from parent giving permission to
upgrade the node's coherence state for address $a$ to $x$\\
\two{DataFromP}{a}{d} & Data $d$ of the cache line received from parent for
address $a$\\
\hline
\multicolumn{2}{|c|}{Outgoing messages}\\
\hline
\three{ReqToC}{c}{a}{x} & Request to child $c$ to downgrade $c$'s
coherence state for address $a$ to $x$\\
\two{ReqToP}{a}{x} & Request to parent to upgrade node's coherence
state for address $a$ to $x$\\
\three{RespToC}{c}{a}{x} & Response to child $c$ granting permission to
upgrade $c'$ coherence state for address $a$ to $x$\\
\three{DataToC}{c}{a}{d} & Data $d$ of the cache line sent to the child
$c$ for address $a$\\
\two{RespToP}{a}{x} & Message to parent denoting that the node's
coherence state for address $a$ has changed to $x$\\
\two{DataToP}{c}{d} & Data $d$ of the cache line sent to the parent
for address $a$\\
\hline
\multicolumn{2}{|c|}{Eviction messages}\\
\hline
\One{EvictReq}{a} & Request to evict cache line for address $a$ (generated by
the node itself)\\
\hline
\end{tabular}
\caption{Incoming, outgoing and eviction messages}
\label{table:messages}
\end{figure}

@@

\begin{figure}
\begin{minipage}{\columnwidth}
\begin{enumerate}
\item A request or a response should not be handled before handling an earlier
response for the same address from the same source

@@

Each message coming from a particular child or the parent is independent of the
messages coming from another source, in addition to being independent of the
messages for a different address. The node can process messages coming from
different sources, and for different addresses in any order, subject to the
ordering requirements in Figure~\ref{list:messageOrder}. We later show that is
easy to enforce this ordering even if the underlying network doesn't enforce
point-to-point FIFO ordering or priorities between different kinds of messages.

To allow maximum flexibility of implementation, the node is given separate
access queues for each kind of message (request of response) from each source
and for each address. An implementation of a protocol using our methodology does
not have to keep a separate access queue for each source and each address, but
it must adhere to the blocking restrictions in Figure~\ref{list:messageOrder}.
These restrictions are necessary to avoid deadlocks. The flexibility of the
abstract protocol allows one to design a wide range of blocking policies.

Our abstract protocol also depends on the designer correctly defining the three
functions shown in Figure~\ref{table:functions}. These functions abstract away
the coherence aspect of the cache coherence protocol. Note that $maxCompat$
function can be derived from $isCompat$ function, if $isCompat$ is defined
correctly. Figure~\ref{fig:msiFunc} shows the functions defined for MSI (--
denotes undefined).

\begin{figure}
\begin{tabular}{|c|p{.63\columnwidth}|}
\hline
isCompat(x, y) & Checks if two different caches can have coherence states
$x$ and $y$, respectively, for the same address\\
\hline
maxCompat(x) & Returns a coherence state value $y$ such that, $\forall z \le y,
isCompat(x, z)$ and $\forall w > y, !isCompat(x, w)$\\
\hline
dataXfer(x, y) & Checks if cache line data has to be transferred by a cache to its
parent on downgrading from $x$ to $y$\\
\hline
\end{tabular}
\caption{Functions to abstract away the coherence aspect}
\label{table:functions}
\end{figure}

@@

We now describe the procedures for a node to handle requests and responses. Note
that a message handler would be triggered only if the corresponding message is
present. If the required resources for handling a message are not present (for
example, if the handling of the request from a child requires sending further
requests to its children or the parent, and the send buffers are full), then the
handler for request from a child would not be triggered.

Figure~\ref{alg:handlePReq} specifies how to handle a node's incoming request
from the parent for downgrading the coherence state of address $a$ to $x$. Note
that a request from the parent can be handled successfully only if the node is
not already waiting for a response from a child.

Figure~\ref{alg:handleCReq} specifies how to handle a node's incoming request
from a child for upgrading the child's coherence state for address $a$ to $x$.
Note that a request from a child can be handled successfully only if the node is
not already waiting for a response from the parent or from any child.

Figure~\ref{alg:handlePResp} specifies how to handle a node's incoming response
from the parent for upgrading the coherence state of address $a$ to $x$.

Figure~\ref{alg:handleCResp} specifies how to handle a node's incoming response
from a child for downgrading the child's coherence state for address $a$ to $x$.

Figure~\ref{alg:handleEvict} specifies how to handle an eviction message (in
order to replace the current line in address $a$ with the line for some other
address, which is given by $replacingAddr(a)$).

@@

\begin{figure}
\begin{algorithmic}
\Rule{handlePReq}{\two{ReqFromP}{a}{x}}
\If {$!isPresent(a)$}
  \State remove$(\two{ReqFromP}{a}{x})$
\ElsIf {$\forall c, !isCWait(c, a)$}
  \If {$state(a) \le x$}
     \State remove$(\two{ReqFromP}{a}{x})$
  \ElsIf {$\forall c, dir(c, a) \le x$}
     \State remove$(\two{ReqFromP}{a}{x})$
     \If {dataXfer$(state(a), x)$}
        \State send$(\two{DataToP}{a}{data(a)})$
     \EndIf
     \State $state(a) \assigns x$
     \State send$(\two{RespToP}{a}{x})$
  \Else
     \ForAll{$c$}
        \If {$dir(c, a) > x$}
           \State $isCWait(c, a) \assigns \True$
           \State $cWait(c, a) \assigns x$
           \State send$(\three{ReqToC}{c}{a}{x})$
        \EndIf
     \EndFor
     \State \textcolor{trig-color}{$whoTrigger(a) \assigns Parent$}
  \EndIf
\EndIf
\EndRule
\end{algorithmic}
\caption{Handle request from parent}
\label{alg:handlePReq}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleCReq}{\three{ReqFromC}{c}{a}{x}}
  \If {$isPresent(a)$}
     \If {$!isPWait(a) \AndAnd \forall c, !isCWait(c, a)$}
        \If {$state(a) \ge x \AndAnd$ \\
              \hspace{2.2cm} $\forall c' \neq c, dir(c', a) \le \text{maxCompat}(x)$}
           \State remove$(\three{ReqFromC}{c}{a}{x})$
           \If {$dir(c,a) \neq \Inv)$}
              \State send$(\three{DataToC}{c}{a}{data(a)})$
           \EndIf
           \State $dir(c, a) \assigns x$
           \State send$(\three{RespToC}{c}{a}{x})$
        \Else
           \If {$state(a) < x$}
              \State $isPWait(a) \assigns \True$
              \State send$(\two{ReqToP}{a}{x})$
           \EndIf
           \ForAll {$c' \neq c$}
              \If {$dir(c', a) > \text{maxCompat}(x)$}
                 \State $isCWait(c', a) \assigns \True$
                 \State $cWait(c', a) \assigns \text{maxCompat}(x)$
                 \State send$(\three{ReqToC}{c'}{a}{\text{maxCompat}(x)})$
              \EndIf
           \EndFor
           \State \textcolor{trig-color}{$whoTrigger(a) \assigns Child$}
           \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
        \EndIf
     \EndIf
  \Else
     \State \textbf{let} $a' \assigns \text{getReplace}(a)$
     \If {$!isPWait(a') \AndAnd \forall c', !isCWait(c', a')$}
        \State send$(\One{EvictReq}{a'})$
        \State \textcolor{trig-color}{$whoTrigger(a) \assigns Evict$}
        \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
        \State $replacingAddr(a') \assigns a$
     \EndIf
  \EndIf
\EndRule
\end{algorithmic}
\caption{Handle request from child}
\label{alg:handleCReq}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleEvict}{\One{EvictReq}{a}}
   \If {$!isPresent(a) \OrOr \forall c, dir(c, a) = \Inv$}
      \State remove$(\One{EvictReq}{a})$
      \If {$state(a) \neq \Inv$}
         \State send$(\two{RespToP}{a}{\Inv})$
         \If {dataXfer$(state(a), \Inv)$}
            \State send$(\two{DataToP}{a}{data(a)})$
         \EndIf
      \EndIf
      \State $isPresent(a) \assigns \False$
      \State allocLine$(replacingAddr(a))$
      \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
   \Else
      \If {$\forall c, !isCWait(c, a)$}
         \ForAll{$c$}
            \If {$dir(c, a) > \Inv$}
               \State $isCWait(c, a) \assigns \True$
               \State $cWait(c, a) \assigns \Inv$
               \State send$(\three{ReqToC}{c}{a}{\Inv})$
            \EndIf
         \EndFor
      \EndIf
   \EndIf
\EndRule
\end{algorithmic}
\caption{Evict a line in address $a$}
\label{alg:handleEvict}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handlePResp}{\two{RespFromP}{a}{x}}
  \State remove$(\two{RespFromP}{a}{x})$
  \If {$state(a) \neq \Inv$}
     \State remove$(\two{DataFromP}{a}{d})$
     \State $data(a) = d$
  \EndIf
  \State $state(a) \assigns x$
  \State $isPWait(a) \assigns \False$ 
  \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
\EndRule
\end{algorithmic}
\caption{Handle response from parent}
\label{alg:handlePResp}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleCResp}{\three{RespFromC}{c}{a}{x}}
  \State remove$(\two{RespFromC}{c}{a}{x})$
  \If {dataXfer$(dir(c,a), x)$}
     \State remove$(\two{DataFromC}{c}{a}{d})$
     \State $data(a) = d$
  \EndIf
  \State $dir(c,a) \assigns x$
  \If {$isCWait(c, a) \AndAnd cWait(c, a) >= x$}
    \State $isCWait(c, a) \assigns \False$
  \EndIf
  {\color{trig-color}
  \If {$whoTrigger(a) = Parent$}
     \State \textbf{trigger} handlePReq
  \ElsIf {$whoTrigger(a) = Evict$}
     \State \textbf{trigger} handleEvict
  \ElsIf {$whoTrigger(a) = Child$}
     \State \textbf{trigger} handleCReq($childTrigger(a)$)
  \EndIf}
\EndRule
\end{algorithmic}
\caption{Handle response from child}
\label{alg:handleCResp}
\end{figure}

Function \emph{getReplace(a)} gives the address of the line that can be replaced
in order to accommodate the line in address $a$. Note that if the line selected
for eviction is waiting for a response from the parent or from any child, then
it can not be replaced.

@@

\begin{figure}
\begin{algorithmic}
\Procedure{allocLine}{$a$}
  \State $isPresent(a) \assigns \True$
  \State \textcolor{trig-color}{$whoTrigger(a) \assigns None$}
  \State $state(a) \assigns \Inv$
  \ForAll {$c$}
     \State $dir(c, a) \assigns \Inv$
     \State $isCWait(c, a) \assigns \False$
  \EndFor
  \State $isPWait(a) \assigns \False$
\EndProcedure
\end{algorithmic}
\caption{Allocate a new cache line for address $a$}
\label{alg:allocLine}
\end{figure}

@@

\item After handling a response (or an evict request) the corresponding handler
must trigger the original handler that requested the received response (or made
an eviction request). The $whoTrigger(a)$ keeps track of the original handler
for each address $a$. In case of a $Child$ trigger, the actual child
corresponding to the original handler is kept track of in $childTrigger(a)$, for
address $a$, in which case $handleCReq$ is triggered for the outstanding request
from that child. In the figures, the pseudo-code corresponding to this is grayed out

@@

The L1 caches, \ie the leaf nodes, do not have any children. Instead, they
handle requests from cores, which resembles handling requests from children. A
store request for a word is essentially a request by the core for write
permissions for the corresponding cache line, and a load request for a word is a
request for read permissions. These permissions can be mapped to the values of
the coherence state.

Because of the absence of children any request from the parent or an eviction
request can be handled without producing more requests. As a result, we do not
need a separate handler for eviction requests. We also do not need to store
$whoTrigger$, since the only trigger that produces more requests, and hence has
to wait for response is a request from the core. We omit the pseudo-code for
handling a request from the core because it is very similar to an instance of
Figure~\ref{alg:handleCReq} with no children.

>>>>

\newcommand{\One}[2]{\text{#1$\langle#2\rangle$}}
\newcommand{\qOne}[2]{\text{#1[$#2$]}}
\newcommand{\qTwo}[3]{\text{#1[$#2$][$#3$}}
\newcommand{\two}[3]{\text{#1$\langle#2,#3\rangle$}}
\newcommand{\three}[4]{\text{#1$\langle#2,#3,#4\rangle$}}
\newcommand{\four}[5]{\text{#1$\langle#2,#3,#4,#5\rangle$}}
\newcommand{\True}{\texttt{True}}
\newcommand{\False}{\texttt{False}}
\newcommand{\assigns}{:=}
\newcommand{\AndAnd}{\;\&\&\;}
\newcommand{\OrOr}{\;||\;}
\newcommand{\Inv}{\text{$Inv$}\;}

@@

A node keeps information about the coherence state of each cache line in the
node as well as in each of its children, \ie the directory.
Figure~\ref{table:states} shows the precise information that a node keeps for
each cache line in the node. The notation we adopt is primarily for conciseness
of description. For instance, the directory need not contain the exact copy of
each of the child's coherence state for each cache line; instead the encoding
can be more efficient. All the information (except $present, data, state$ and
$dir$) corresponds to a request being handled for a given address; an
implementation can store this information only for the requests being processed
instead of for each address.

\begin{figure}
\begin{tabular}{|l|p{.64\columnwidth}|}
\hline
isPresent[a] & Is cache line for address $a$ present?\\
\hline
data[a] & Cache line data for address $a$\\
\hline
state[a] & Coherence state for address $a$\\
\hline
dir[a][c] & Directory's copy of the coherence state for address $a$ in child
$c$\\
\hline
isCWait[a] & Is node waiting for a response from any child for address $a$?\\
\hline
cWaitState[a] & The coherence state to which the node has requested the children to
downgrade to, for address $a$\\
\hline
isPWait[a] & Is node waiting for a response from its parent for
address $a$?\\
\hline
isPReq[a] & Is node handling a request from parent for addess $a$\\
\hline
isCReq[a] & Is node handling a request from a child for address $a$\\
\hline
whichCReq[a] & The child whose request the node is handling for address $a$\\
\hline
isEvict[a] & Is the cache line for address $a$ being evicted?\\
\hline
replacingAddr[a] & If the cache line for address $a$ is being evicted, the address replacing the line\\
\hline
\end{tabular}
\caption{Information that the node keeps for a cache line}
\label{table:states}
\end{figure}

The values of the coherence state (and hence the directory state) belong to a
totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
defined between any two set elements and this relation is
transitive. For instance, in an MSI protocol, the values of the coherence state
are from the set $\{M, S, I\}$ where $I < S$, $S < M$ and $I < M$. Our abstract
protocol does not assign any semantics or meaning to the values of the coherence
states. In a concrete cache coherence protocol, the values of the states are
associated with permissions for reading and writing, and the state with fewer
permissions is $<$ the state with more permissions. We use \Inv to denote the
least element in the set containing the values of the coherence state of a line.
Typically, this denotes that the node has no permissions for reading or writing
that line.

@@

Figure~\ref{table:messages} shows the incoming and outgoing messages to and from
the node, respectively.

\begin{figure}
\begin{tabular}{|l|p{.6\columnwidth}|}
\hline
\multicolumn{2}{|c|}{Incoming messages}\\
\hline
\three{ReqFromC}{c}{a}{x} & Request from child $c$ upgrade $c$'s coherence
state for address $a$ to $x$\\
\two{ReqFromP}{a}{x} & Request from parent to downgrade node's coherence
state for address $a$ to $x$\\
\three{RespFromC}{c}{a}{x} & Message from child $c$ denoting that $c$'s
coherence state for address $a$ has downgraded to $x$\\
\three{DataFromC}{c}{a}{d} & Data $d$ of the cache line received from child $c$
for address $a$\\
\two{RespFromP}{a}{x} & Response from parent giving permission to
upgrade the node's coherence state for address $a$ to $x$\\
\two{DataFromP}{a}{d} & Data $d$ of the cache line received from parent for
address $a$\\
\hline
\multicolumn{2}{|c|}{Outgoing messages}\\
\hline
\three{ReqToC}{c}{a}{x} & Request to child $c$ to downgrade $c$'s
coherence state for address $a$ to $x$\\
\two{ReqToP}{a}{x} & Request to parent to upgrade node's coherence
state for address $a$ to $x$\\
\three{RespToC}{c}{a}{x} & Response to child $c$ granting permission to
upgrade $c'$ coherence state for address $a$ to $x$\\
\three{DataToC}{c}{a}{d} & Data $d$ of the cache line sent to the child
$c$ for address $a$\\
\two{RespToP}{a}{x} & Message to parent denoting that the node's
coherence state for address $a$ has downgraded to $x$\\
\two{DataToP}{c}{d} & Data $d$ of the cache line sent to the parent
for address $a$\\
\hline
\end{tabular}
\caption{Incoming and outgoing messages}
\label{table:messages}
\end{figure}

Each message coming from a particular child or the parent is independent of the
messages coming from another source, in addition to being independent of the
messages for a different address. The node can process messages coming from
different sources, or corresponding to different addresses, in any order,
subject to the ordering requirements in Figure~\ref{list:messageOrder}. The
message ordering restrictions are necessary to avoid deadlocks. We later show
that is easy to enforce this ordering even if the underlying network doesn't
enforce point-to-point FIFO ordering or priorities between different kinds of
messages.

@@

\begin{figure}
\begin{minipage}{\columnwidth}
\begin{enumerate}
\item A request or a response should not be handled before handling a response
sent earlier from the same source for the same address

@@

%To allow maximum flexibility of implementation, the node is given separate
%access queues for each kind of message (request of response) from each source
%and for each address (\ie \qTwo{ReqFromC}{c}{a}, \qOne{ReqFromPQ}{a},
%\qTwo{RespFromCQ}{c}{a}, \qTwo{ReqToCQ}{c}{a}, \qOne{RespToPQ}{a}, \etc). An
%implementation of a protocol using our methodology does not have to keep a
%separate access queue for each source and each address, but it must adhere to
%the blocking restrictions in Figure~\ref{list:messageOrder}.  These
%restrictions are necessary to avoid deadlocks. The flexibility of the abstract
%protocol allows one to design a wide range of blocking policies.

Our abstract protocol also depends on the designer correctly defining the three
functions shown in Figure~\ref{table:functions}. These functions abstract away
the coherence aspect of the cache coherence protocol. Note that $maxCompat$
function can be derived from $isCompat$ function, if $isCompat$ is defined
correctly. Figure~\ref{fig:msiFunc} shows the functions defined for MSI (``--''
denotes undefined).

\begin{figure}
\begin{tabular}{|c|p{.63\columnwidth}|}
\hline
isCompat(x, y) & Returns if two different sibling caches can have coherence
states $x$ and $y$, respectively, for the same address or not\\
\hline
maxCompat(x) & Returns a coherence state value $y$ such that, $\forall z \le y,
isCompat(x, z)$ and $\forall w > y, !isCompat(x, w)$\\
\hline
dataXfer(x, y) & Return if cache line data has to be transferred by a cache to its
parent on downgrading ist coherence state from $x$ to $y$ or not\\
\hline
\end{tabular}
\caption{Functions to abstract away the coherence aspect}
\label{table:functions}
\end{figure}

@@

We now describe the how a node handles requests and responses.

Note
that a message handler would be triggered only if the corresponding message is
present. If the required resources for handling a message are not present (for
example, if the handling of the request requires sending further requests to
its children or the parent, and the corresponding send buffers are full), then
the handler for that message would not be triggered.

Figure~\ref{alg:handlePReq} specifies how a node handles an incoming request
from the parent for downgrading the coherence state of address $a$ to $x$. Note
that a request from the parent can be handled successfully only if the node is
not already waiting for a response from some child.

Figure~\ref{alg:handleCReq} specifies how a node handles an incoming request
from a child for upgrading the child's coherence state for address $a$ to $x$.
Note that a request from a child can be handled successfully only if the node
is not already waiting for a response from the parent or from some child.

Figure~\ref{alg:handlePResp} specifies how a node handles an incoming response
from the parent for permitting upgrade of the coherence state of address $a$ to
$x$.

Figure~\ref{alg:handleCResp} specifies how a node handles an incoming response
from a child notifying the downgrade of the child's coherence state for address
$a$ to $x$.

XXX Figure~\ref{alg:handleEvict} specifies how a cache line replacement is handled.

@@

\begin{figure}
\begin{algorithmic}
\Rule{handlePReq}{a}
   \State let mesg \two{ReqFromP}{a}{x} exist in \qOne{ReqFromPQ}{a}
   \If {$\forall c, !isCWait(c][a)$}
     \If {$state(a) \le x$}
        \State remove$(\two{ReqFromP}{a}{x})$
     \State remove$(\two{ReqFromP}{a}{x})$
     \ElsIf {$\forall c, dir(c][a) \le x$}
        \State remove$(\two{ReqFromP}{a}{x})$
        \If {dataXfer$(state(a), x)$}
           \State send$(\two{DataToP}{a}{data(a)})$
        \EndIf
        \State $state(a) \assigns x$
        \State send$(\two{RespToP}{a}{x})$
     \Else
        \ForAll{$c$}
           \If {$dir(c][a) > x$}
              \State $isCWait(c][a) \assigns \True$
              \State $cWait(c][a) \assigns x$
              \State send$(\three{ReqToC}{c}{a}{x})$
           \EndIf
        \EndFor
        \State \textcolor{trig-color}{$whoTrigger(a) \assigns Parent$}
     \EndIf
   \EndIf
\EndRule
\end{algorithmic}
\caption{Handle request from parent}
\label{alg:handlePReq}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleCReq}{$c, a$}
   \State let mesg \two{ReqFromC}{c}{a}{x} exist in \qTwo{ReqFromCQ}{c}{a}
   \If {$isPresent(a)$}
      \If {$!isPWait(a) \AndAnd \forall c, !isCWait(c][a)$}
         \If {$state(a) \ge x \AndAnd$ \\
               \hspace{2.2cm} $\forall c' \neq c, dir(c', a) \le \text{maxCompat}(x)$}
            \State remove$(\three{ReqFromC}{c}{a}{x})$
            \If {$dir(c,a) \neq \Inv)$}
               \State send$(\three{DataToC}{c}{a}{data(a)})$
            \EndIf
            \State $dir(c][a) \assigns x$
            \State send$(\three{RespToC}{c}{a}{x})$
         \Else
            \If {$state(a) < x$}
               \State $isPWait(a) \assigns \True$
               \State send$(\two{ReqToP}{a}{x})$
            \EndIf
            \ForAll {$c' \neq c$}
               \If {$dir(c', a) > \text{maxCompat}(x)$}
                  \State $isCWait(c', a) \assigns \True$
                  \State $cWait(c', a) \assigns \text{maxCompat}(x)$
                  \State send$(\three{ReqToC}{c'}{a}{\text{maxCompat}(x)})$
               \EndIf
            \EndFor
            \State \textcolor{trig-color}{$whoTrigger(a) \assigns Child$}
            \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
         \EndIf
      \EndIf
   \Else
      \State \textbf{let} $a' \assigns \text{getReplace}(a)$
      \If {$!isPWait(a') \AndAnd \forall c', !isCWait(c', a')$}
         \State \textcolor{trig-color}{$whoTrigger(a) \assigns Evict$}
         \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
         \State $replacingAddr(a') \assigns a$
         \State \textcolor{trig-color}{\textbf{trigger} handleEvict($a$)}
      \EndIf
   \EndIf
\EndRule
\end{algorithmic}
\caption{Handle request from child}
\label{alg:handleCReq}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleEvict}{$a$}
   \If {$!isPresent(a) \OrOr \forall c, dir(c][a) = \Inv$}
      \If {$state(a) \neq \Inv$}
         \State send$(\two{RespToP}{a}{\Inv})$
         \If {dataXfer$(state(a), \Inv)$}
            \State send$(\two{DataToP}{a}{data(a)})$
         \EndIf
      \EndIf
      \State $isPresent(a) \assigns \False$
      \State allocLine$(replacingAddr(a))$
      \State \textcolor{trig-color}{$whoTrigger(a) \assigns None$}
      \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a), a$)}
   \Else
      \If {$\forall c, !isCWait(c][a)$}
         \ForAll{$c$}
            \If {$dir(c][a) > \Inv$}
               \State $isCWait(c][a) \assigns \True$
               \State $cWait(c][a) \assigns \Inv$
               \State send$(\three{ReqToC}{c}{a}{\Inv})$
            \EndIf
         \EndFor
      \EndIf
   \EndIf
\EndRule
\end{algorithmic}
\caption{Evict a line in address $a$}
\label{alg:handleEvict}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handlePResp}{$a$}
   \State let mesg \two{RespFromP}{a}{x} exist in \qOne{RespFromPQ}{a}
   \State remove$(\two{RespFromP}{a}{x})$
   \If {$state(a) \neq \Inv$}
      \State let mesg \two{DataFromP}{a}{d} exist in \qOne{DataFromPQ}{a}
      \State remove$(\two{DataFromP}{a}{d})$
      \State $data(a) = d$
   \EndIf
   \State $state(a) \assigns x$
   \State $isPWait(a) \assigns \False$ 
   \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a), a$)}
\EndRule
\end{algorithmic}
\caption{Handle response from parent}
\label{alg:handlePResp}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleCResp}{$c, a$}
   \State let mesg \three{RespFromC}{c}{a}{x} exist in \qTwo{RespFromCQ}{c}{a}
   \State remove$(\three{RespFromC}{c}{a}{x})$
   \If {dataXfer$(dir(c,a), x)$}
      \State let mesg \three{DataFromC}{c}{a}{d} exist in \qTwo{DataFromPQ}{c}{a}
      \State remove$(\three{DataFromC}{c}{a}{d})$
      \State $data(a) = d$
   \EndIf
   \State $dir(c,a) \assigns x$
   \If {$isCWait(c][a) \AndAnd cWait(c][a) >= x$}
     \State $isCWait(c][a) \assigns \False$
   \EndIf
   {\color{trig-color}
   \If {$whoTrigger(a) = Parent$}
      \State \textbf{trigger} handlePReq($a$)
   \ElsIf {$whoTrigger(a) = Evict$}
      \State \textbf{trigger} handleEvict($(a$)
   \ElsIf {$whoTrigger(a) = Child$}
      \State \textbf{trigger} handleCReq($childTrigger(a), a$)
   \EndIf}
\EndRule
\end{algorithmic}
\caption{Handle response from child}
\label{alg:handleCResp}
\end{figure}

Function \emph{getReplace(a)}, used in \textsc{handleCReq} gives the address of
the line that can be replaced in order to accommodate the line in address $a$.
Note that if the line selected for eviction is waiting for a response from the
parent or from any child, then it can not be replaced.

@@

\begin{figure}
\begin{algorithmic}
\Procedure{allocLine}{$a$}
  \State $isPresent(a) \assigns \True$
  \State \textcolor{trig-color}{$whoTrigger(a) \assigns None$}
  \State $state(a) \assigns \Inv$
  \ForAll {$c$}
     \State $dir(c][a) \assigns \Inv$
     \State $isCWait(c][a) \assigns \False$
  \EndFor
  \State $isPWait(a) \assigns \False$
\EndProcedure
\end{algorithmic}
\caption{Allocate a new cache line for address $a$}
\label{alg:allocLine}
\end{figure}

@@

\item After handling a response (or an eviction) the corresponding handler
must trigger the original handler that requested the received response (or made
an eviction request). The $whoTrigger(a)$ keeps track of the original handler
for each address $a$. In case of a $Child$ trigger, the actual child
corresponding to the original handler is kept track of in $childTrigger(a)$, for
address $a$, in which case $handleCReq$ is triggered for the outstanding request
from that child. In the figures, the pseudo-code corresponding to this is grayed out

@@

The L1 caches, \ie the leaf nodes, do not have any children. Instead, they
handle requests from cores. A store word request is essentially a request by
the core for write permissions for the corresponding cache line, and a load
word request is a request for read permissions. These permissions can be mapped
to the values of the coherence state. Because of the absence of children any
request from the parent as well as any eviction can be handled without
producing more requests. As a result, we do not need a separate handler for
eviction requests. We also do not need to store $whoTrigger(a)$, since the only
trigger that produces more requests, and hence has to wait for response is a
request from the core. We omit the pseudo-code for handling a request from the
core because it is very similar to an instance of \textsc{handleCReq} with no
children.
